Building and Checking a Suffix Array by Induced Sorting

### 2017-08-11 new-draft/draft.tex
2017/08/11

In the accepted paper, the LCP-value $\ell$ of ${\sf suf}(i)$ and ${\sf suf}(j)$ is assumed to be right if ${\sf sub}(i, i + \ell - 1) = {\sf sub}(j, j + \ell - 1)$ and $x[i + \ell] \ne x[j + \ell]$. Given that $\ell$ is correct, the lexical order of ${\sf suf}(i)$ and ${\sf suf}(j)$ is checked by comparing $x[i + \ell]$ and $x[j + \ell]$.

@@

While the research on developing various SA builders using the IS method is involving, the developed software are becoming complex and appear as open-source without guarantee. A constructed SA should be checked to detect potential errors caused by implementation bugs and other malfunctions. The software packages for DC3 and eSAIS provide a checker based on the idea presented in~\cite{Dementiev2008a}. When running on external-memory model, the cost taken  by this checker is rather high as it performs two passes of integer sorts for arranging $\mathcal{O}(n)$ fixed-size tuples. In this paper, we propose two methods to enable the IS method to build and check an SA simultaneously. In particular, the first method can be extended to check both suffix and LCP arrays. We employ the Karp-Rabin fingerprinting technique~\cite{Karp1987} to design two probabilistic algorithms, in terms of that their checking results are wrong with a negligible probability. For analysis, we first use new substring sorting and naming methods to improve the design of DSA-IS and then combine the second checking method into the adapted DSA-IS for evaluating the checking overhead. Our experimental results indicate that the time, space and I/O volume for checking SA is considerably smaller than that for building SA.

@@

Partition $x$ into multiple S*-type substrings, each substring contains two S*-type characters and any two neighboring substrings overlap an S*-type character. We produce a reduced string $x_1$ by replacing each substring with its name, where the name indicates the rank of the	 substring among all. In this paper, we use "rank" and "name" interchangeably to represent the lexical order of substrings.

@@

	\item Split $x$ into blocks and sort substrings/suffixes of each block by calling SA-IS. The heading characters in need are copied to external memory in their access order. \label{dsais_sorting_method:1}

	\item Sort the substrings/suffixes by their heading characters and the ranks of their successors in an external-memory heap. \label{dsais_sorting_method:2}
\end{enumerate}

@@

\begin{enumerate}[S1']
	\item Sort the long in each block, copy the heading characters in need to external memory by S\ref{dsais_sorting_method:1}. Meanwhile, sort the short in each block, copy the short to external memory in their sorted order.~\label{dsaism_sorting_method:1}

	\item Sort the long in $x$ by S\ref{dsais_sorting_method:2}, copy the leftmost $D$ characters of the long to external memory in their sorted order.~\label{dsaism_sorting_method:2}

@@

The sorted short S*-type substrings in each block are organized as a sequence in external memory after S\ref{dsaism_sorting_method:1}', while the sorted long S*-type substrings in $x$ are also organized as a sequence in external memory after S\ref{dsaism_sorting_method:2}'. Assume $x$ is split into $k$ blocks, the multi-way sorter in S\ref{dsaism_sorting_method:3}' maintains an internal-memory heap to determine the lexical order of substrings belonging to different sequences. The heap contains at most $k + 1$ substrings at any point of time. It compares any two substring in $\mathcal{O}(D)$ time by conducting a literal string comparison\footnote{If the leftmost $D$ characters of a long S*-type substring is equal to a short S*-type substring, then the short is lexicographically greater than the long.}.

The above sorting method has a good performance if $D$ is small and the majority of S*-type substrings in $x$ are short. This is commonly satisfied in real-world datasets. We next describe a method for naming the S*-type substrings when they are being sorted according to S\ref{dsaism_sorting_method:1}'-S\ref{dsaism_sorting_method:3}'. The key point is to check equality of two suffixes successively popped from the heap. For this, we literally compare the two substrings in $\mathcal{O}(D)$ time if either of them is short; otherwise, we compare their names in $\mathcal{O}(1)$ time instead. Recall that the rank of each long S*-type substring is determined when inducing its predecessor, this value is reused to represent its name herein\footnote{A similar technique has been used to merge the substring sorting and naming process in SAIS-PQ.}. The experimental results in Section~\ref{sec:experiments} demonstrate that, by using the new substring sorting and naming methods, the adapted DSA-IS, called DSA-IS+, only takes half as much disk space as eSAIS.

@@

A running example of S\ref{induction_phase:2}''-S\ref{induction_phase:4}'' is shown in Fig.~\ref{fig:example1}. Given $sa^*$ is already known, line 6 inserts each S*-type suffix into the corresponding bucket. For example, ${\sf suf}(2)$, ${\sf suf}(5)$, ${\sf suf}(8)$ are sequentially placed into the current rightmost empty position of ${\sf sa\_bkt}(i)$, where the insertion order corresponds to their sorted order indicated by $sa^*$. The next step is to find the leftmost position of each bucket (marked by the symbol $\wedge$) and scan $sa$ rightward for inducing the order of L-type suffixes. For this, we first check $sa[0] = 14$ (marked by the symbol @) and induce the predecessor of ${\sf suf}(14)$ in lines 8-9. Because $x[13] = i$ is L-type, we put ${\sf suf}(13)$ into the current leftmost empty position in ${\sf sa\_bkt_L}(i)$. To step through $sa$ in this way, we get all the L-type suffixes sorted in line 17. Afterward, we find the rightmost position of each bucket and scan $sa$ leftward for inducing the order of S-type suffixes. When scanning $sa[14] = 4$ in lines 19-20, we see $x[3] = i$ is S-type and thus put ${\sf suf}(3)$ into the current rightmost empty position in ${\sf sa\_bkt_S}(i)$. Following the same idea, we get all the S-type suffixes sorted in line 28. Following the discussion, we show in Lemma~\ref{lemma:2} a set of sufficient and necessary conditions for checking SA.

@@

\begin{lemma} \label{lemma:2}
    For any IS suffix sorting algorithm, its output $sa[0, n)$ is the SA for $x[0, n)$ if and only if the following conditions are satisfied at the top recursion level: \\
    (1) $sa^*$ is correctly computed. \\
    (2) S\ref{induction_phase:1}''-S\ref{induction_phase:4}'' are correctly implemented. \\
\end{lemma}

@@

Further, suppose $sa[i] = sa[j]$ and $i \ne j$, then $sa[p] = sa[q]$ and $p \ne q$. Repeating this reasoning process by replacing $(i, j)$ with $(p, q)$ until ${\sf suf}(sa[i])$ is S*-type, then $sa^*$ must contain duplicate elements. However, each element in $sa^*$ is unique as $sa^*$ is corrected computed. This leads to a contradiction.

@@

It was reported in~\cite{Fischer11, Bingmann12} that the IS method can be used to construct LCP array as well. For any two neighboring suffixes in $sa$, their LCP-value is zero if they start with different characters; otherwise, the LCP-value of them is one greater than that of their successors. This implies the possibility for extending Method A to check both suffix and LCP arrays. We design Algorithm~\ref{alg:2} to check $sa$ and $lcp$ when they are induced from $sa^*$ and $lcp^*$ at the top recursion level of an IS builder. In this algorithm, ${\sf sa_{I1}}(c)$ and ${\sf sa_{I2}}(c)$ are two sequences respectively induced into ${\sf sa\_bkt_L}(c)$ and ${\sf sa\_bkt_S}(c)$, while ${\sf sa_{S1}}(c)$ and ${\sf sa_{S2}}(c)$ are two sequences respectively scanned from ${\sf sa\_bkt_L}(c)$ and ${\sf sa\_bkt_S}(c)$. These values are used to check the second condition of Lemma~\ref{lemma:3} later. We also compute ${\sf lcp_{I1}}(c)/{\sf lcp_{I2}}(c)$ and ${\sf lcp_{S1}}(c)/{\sf lcp_{S2}}(c)$ and check their equality in the meantime.

@@

        Check if ${\sf sa_{I2}}(c) = {\sf sa_{S2}}(c)$, ${\sf lcp_{I2}}(c) = {\sf lcp_{S2}}(c)$. \\
	}
	\end{algorithm*}

@@

Assume the induction phase at the top recursion level is correctly implemented, Algorithm~\ref{alg:2} employs the method proposed in~\cite{wu2017} to ensure the correctness of $sa^*$ and $lcp^*$. To be specific, for two neighboring suffixes ${\sf suf}(i)$ and ${\sf suf}(j)$ in $sa^*$, their LCP-value $\ell$ in $lcp^*$ is right if ${\sf sub}(i, i + \ell - 1) = {\sf sub}(j, j + \ell - 1)$ and $x[i + \ell] \ne x[j + \ell]$, where the equality of ${\sf sub}(i, i + \ell - 1)$ and ${\sf sub}(j, j + \ell - 1)$ is checked by comparing the fingerprints of these substrings. Given that $\ell$ is correct, the lexical order of ${\sf suf}(i)$ and ${\sf suf}(j)$ is checked by comparing $x[i + \ell]$ with $x[j + \ell]$. This method computes the fingerprints of all the target substrings within a sorting complexity proportional to $n_1$. Because at most one out of every two suffixes is S*-type, the checking overhead is much less than building consumption. By contrast, Algorithm~\ref{alg:3} takes overall $\mathcal{O}(n)$ to compute all the required fingerprints.

@@

By assuming the induction phase at the top recursion level is correctly implemented, we proposed two methods that enable any IS suffix sorter to build and check an SA simultaneously. The probabilistic algorithm designed by the second method is rather lightweight, it takes negligible time and space compared with the existing IS suffix sorting and checking algorithms. We also made our first attempt to improve the space performance of DSA-IS using new substring and naming methods. Our experimental results show that the program for the adapted algorithm DSA-IS+ runs as fast as that for eSAIS and consumes only half as much disk space as the latter on various real-world datasets. We are now designing and implementing a novel IS suffix sorter that takes no more than 1n work space on external memory model. Theoretically, this suffix sorter has a better space performance than fSAIS under the same circumstances.

>>>>

The LCP-value $\ell$ of ${\sf suf}(i)$ and ${\sf suf}(j)$ is right if ${\sf sub}(i, i + \ell - 1) = {\sf sub}(j, j + \ell - 1)$ and $x[i + \ell] \ne x[j + \ell]$. In the accepted paper, we check equality of ${\sf sub}(i, i + \ell - 1)$ and ${\sf sub}(j, j + \ell - 1)$ by comparing their fingerprints. Given that $\ell$ is correct, the lexical order of ${\sf suf}(i)$ and ${\sf suf}(j)$ is determined by comparing $x[i + \ell]$ and $x[j + \ell]$.

@@

While the research on developing various SA builders using the IS method is evolving, the developed software are becoming complex and appear as open-source without guarantee. A constructed SA should be checked to detect potential errors caused by implementation bugs and other malfunctions. The software packages for DC3 and eSAIS provide a checker based on the idea presented in~\cite{Dementiev2008a}. When running on external-memory model, the cost taken  by this checker is rather high as it performs two passes of integer sorts for arranging $\mathcal{O}(n)$ fixed-size tuples. In this paper, we propose two methods to enable the IS method to build and check an SA simultaneously. In particular, the first method can be extended to check both suffix and LCP arrays. We employ the Karp-Rabin fingerprinting technique~\cite{Karp1987} to design two probabilistic algorithms, in terms of that their checking results are wrong with a negligible probability. For analysis, we first use new substring sorting and naming methods to improve the design of DSA-IS and then combine the second checking method into the adapted DSA-IS for evaluating the checking overhead. Our experimental results indicate that the time, space and I/O volume for checking SA is considerably smaller than that for building SA.

@@

Partition $x$ into multiple S*-type substrings, each substring contains two S*-type characters and any two neighboring substrings overlap an S*-type character. We produce a reduced string $x_1$ by replacing each substring with its name, where the name represents the rank of this	 substring among all. In this paper, we use "rank" and "name" interchangeably to indicate the lexical order of substrings.

@@

	\item Split $x$ into blocks and sort substrings/suffixes in each block by calling SA-IS. The heading characters in need are copied to external memory in their access order. \label{dsais_sorting_method:1}

	\item Sort the substrings/suffixes in $x$ by their heading characters and the ranks of their successors in an external-memory heap. \label{dsais_sorting_method:2}
\end{enumerate}

@@

\begin{enumerate}[S1']
	\item Reuse S\ref{dsais_sorting_method:1} to sort the short and the long in each block. Copy the heading characters for sorting the long to external memory in their access order and copy the short to external memory in their sorted order.~\label{dsaism_sorting_method:1}

	\item Reuse S\ref{dsais_sorting_method:2} to sort the long in $x$. Copy the leftmost $D$ characters of the long to external memory in their sorted order.~\label{dsaism_sorting_method:2}

@@

The short S*-type substrings in each block are already sorted and organized as a sequence in external memory after S\ref{dsaism_sorting_method:1}', while the long S*-type substrings in $x$ are already sorted and organized as a sequence in external memory after S\ref{dsaism_sorting_method:2}'. Assume $x$ is split into $k$ blocks, the multi-way sorter in S\ref{dsaism_sorting_method:3}' maintains an internal-memory heap to determine the lexical order of substrings belonging to different sequences. The heap contains at most $k + 1$ substrings at any point of time. It performs a literal string comparison to compare any two substrings in $\mathcal{O}(D)$\footnote{If the leftmost $D$ characters of a long S*-type substring is equal to a short S*-type substring, then the short is lexicographically greater than the long.}.

The above sorting method has a good performance if $D$ is small and the majority of S*-type substrings in $x$ are short. This is commonly satisfied in real-world datasets. We next describe a method for naming the S*-type substrings when they are being sorted according to S\ref{dsaism_sorting_method:1}'-S\ref{dsaism_sorting_method:3}'. The key point is to check equality of two substrings successively popped from the heap. For this, we literally compare the two substrings in $\mathcal{O}(D)$ time if either of them is short; otherwise, we compare their names in $\mathcal{O}(1)$ time instead. Recall that the rank of each long S*-type substring is determined when inducing its predecessor, this value is reused to represent its name herein\footnote{A similar technique has been used to merge the substring sorting and naming processes in SAIS-PQ.}. Our experiments in Section~\ref{sec:experiments} show that, by using the new substring sorting and naming methods, the adapted DSA-IS, called DSA-IS+, only takes half as much disk space as eSAIS.

@@

A running example of S\ref{induction_phase:2}''-S\ref{induction_phase:4}'' is shown in Fig.~\ref{fig:example1}. Given $sa^*$ is already known, line 6 inserts each S*-type suffix into the corresponding bucket. For example, ${\sf suf}(2)$, ${\sf suf}(5)$, ${\sf suf}(8)$ are sequentially placed into the current rightmost empty position of ${\sf sa\_bkt}(i)$, where the insertion order corresponds to their sorted order indicated by $sa^*$. The next step is to find the leftmost position of each bucket (marked by the symbol $\wedge$) and scan $sa$ rightward for inducing the order of L-type suffixes. For this, we first check $sa[0] = 14$ (marked by the symbol @) and induce the predecessor of ${\sf suf}(14)$ in lines 8-9. Because $x[13] = i$ is L-type, we put ${\sf suf}(13)$ into the current leftmost empty position in ${\sf sa\_bkt_L}(i)$. To step through $sa$ in this way, we get all the L-type suffixes sorted in line 17. Afterward, we find the rightmost position of each bucket and scan $sa$ leftward for inducing the order of S-type suffixes. When scanning $sa[14] = 4$ in lines 19-20, we see $x[3] = i$ is S-type and thus put ${\sf suf}(3)$ into the current rightmost empty position in ${\sf sa\_bkt_S}(i)$. Following the same idea, we get all the S-type suffixes sorted in line 28. Following the discussion, we show in Lemma~\ref{lemma:2} a set of sufficient conditions for checking SA.

@@

\begin{lemma} \label{lemma:2}
    For any IS suffix sorter, its output $sa[0, n)$ is the SA for $x[0, n)$ if the following conditions are satisfied at the top recursion level: \\
    (1) $sa^*$ is correctly computed. \\
    (2) S\ref{induction_phase:1}''-S\ref{induction_phase:4}'' are correctly implemented. \\
\end{lemma}

@@

Suppose $sa[i] = sa[j]$ and $i \ne j$, because S\ref{induction_phase:1}''-S\ref{induction_phase:4}'' are correctly implemented, $sa[p] = sa[q]$ and $p \ne q$. Repeating this reasoning process by replacing $(i, j)$ with $(p, q)$ until ${\sf suf}(sa[i])$ is S*-type, then $sa^*$ must contain duplicate elements. However, each element in $sa^*$ is unique as $sa^*$ is correctly computed, leading to a contradiction. This satisfies the first condition of Lemma~\ref{lemma:1}.

@@

It was reported in~\cite{Fischer11} and~\cite{Bingmann12} that the IS method can be used to construct LCP array as well. For any two neighboring suffixes in $sa$, their LCP-value is zero if they start with different characters; otherwise, the LCP-value of them is one greater than that of their successors. This implies the possibility for extending Method A to check both suffix and LCP arrays. We design Algorithm~\ref{alg:2} to check $sa$ and $lcp$ when they are induced from $sa^*$ and $lcp^*$ at the top recursion level of an IS builder. In this algorithm, ${\sf sa_{I1}}(c)$ and ${\sf sa_{I2}}(c)$ are two sequences respectively induced into ${\sf sa\_bkt_L}(c)$ and ${\sf sa\_bkt_S}(c)$, while ${\sf sa_{S1}}(c)$ and ${\sf sa_{S2}}(c)$ are two sequences respectively scanned from ${\sf sa\_bkt_L}(c)$ and ${\sf sa\_bkt_S}(c)$. These values are compared for checking the second condition of Lemma~\ref{lemma:3}. The algorithm also computes and compares ${\sf lcp_{I1}}(c)/{\sf lcp_{I2}}(c)$ and ${\sf lcp_{S1}}(c)/{\sf lcp_{S2}}(c)$ in the meantime.

@@

        Check if ${\sf sa_{I2}}(c) = {\sf sa_{S2}}(c)$ and ${\sf lcp_{I2}}(c) = {\sf lcp_{S2}}(c)$. \\
	}
	\end{algorithm*}

@@

A key operation of Algorithms~\ref{alg:2} and~\ref{alg:3} is to check equality of two integer sequences by comparing their fingerprints. This is done in linear time and constant RAM space by using the Karp-Rabin fingerprinting technique. Algorithm~\ref{alg:2} also employs the method presented in~\cite{wu2017} to verify $sa^*$ and $lcp^*$. For any two neighboring suffixes in $sa^*$, this method checks their lexical order indicated by $sa^*$ and the LCP-value $\ell$ indicated by $lcp^*$ as follows: (1) check if the fingerprints of their leftmost $\ell$ characters are equal; (2) check if their $(\ell + 1)$-th characters are different; (3) check if their lexical order corresponds to that of their $(\ell + 1)$-th characters. It should be noticed that this method checks all the neighboring suffix pairs in batch within a sorting complexity. Because at most one out of every two suffixes is S*-type and the checking process is only executed at the top recursion level, the consumption for checking $sa^*$ and $lcp^*$ is much less than that for building the two arrays by the IS method.

@@

By assuming the induction phase at the top recursion level is correctly implemented, we proposed two methods that enable any IS suffix sorter to build and check an SA simultaneously. The probabilistic algorithm designed by the second method is rather lightweight, it takes negligible time and space compared with the existing IS suffix sorting and checking algorithms. In this paper, we also made the first attempt to improve the performance of DSA-IS using new substring and naming methods. Our program for the adapted algorithm DSA-IS+ runs as fast as that for eSAIS and consumes only half as much disk space as the latter on various real-world datasets. We are now designing and implementing a novel IS suffix sorter that takes no more than 1n work space on external memory model. Theoretically, this suffix sorter has a better space performance than fSAIS under the same circumstances.
